---
title: "p8105_hw2_yc4384"
author: "Yangyang Chen"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
```

## Problem 1

**First:**

* Clean the data in `pols-month.csv`. 
* Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`.
* Replace month number with month name 
* Create a `president` variable taking values `gop` and `dem`.
* Remove `prez_dem` and `prez_gop`; and remove the `day` variable.
```{r}
pols_df = 
  read_csv("data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into = c("year", "month", "day")) |> 
  mutate(month = month.abb[as.numeric(month)]) |> 
  mutate(president  = gov_dem) |> 
  select(-c(prez_dem,prez_gop)) |> 
  select(-day)
```

**Second:** 

* Clean the data in snp.csv using a similar process to the above.
* Arrange according to year and month.
* Organize so that `year` and `month` are the leading columns.
```{r}
snp_df = 
  read_csv("data/snp.csv") |> 
  janitor::clean_names() |> 
  separate(date, into = c("year", "month", "day")) |> 
  mutate(month = month.abb[as.numeric(month)]) |> 
  select(-day) |> 
  relocate(year, month)
```

**Third:** 

* Tidy the unemployment data so that it can be merged with the previous datasets. 
* This process will involve switching from “wide” to “long” format.
* Ensuring that key variables have the same name.
* Ensuring that key variables take the same values.

```{r}
unemploy_df = 
  read_csv("data/unemployment.csv") |> 
  janitor::clean_names() |> 
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemploy_rate"
  ) |> 
  relocate(year, month)
```

